A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis

### 2022-02-01 fstats_pca_lewontin.tex
rev3

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to analyze human genetic variation. Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I show that admixture corresponds to I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{boca2020}.

@@

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}. These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations. These analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.

@@

\subsection{Formal Definition of $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci. Let $x_{il}$ denote the frequency of an arbitrary allele at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. On trees, all disjoint paths are independent (i.e. orthogonal) from each other, and thus the external branches vanish under the projection.

The geometric approach of \cite{oteo-garcia2021} assumes a very high-dimensional space, as each SNP in a data set adds a dimension, and data sets have commonly a million or more SNPs. Such high-dimensional spaces are hard to visualize and analyze, however, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}. Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.

@@

\subsubsection{Missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}. The theory developed here suggests that developing a PCA-based $F$-statistics framework might be more powerful, as related populations could be used to fill in missing data.

@@

An alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}. These associations are not exact, and impacted e.g. by the size and composition of the analyzed data set. On the other hand, we can use $F$-statistics to formally test orthogonality. Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary. Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e. between-population diversity is independent of within-population diversity). This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages.

>>>>

\title{A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Principal Component Analysis (PCA) and $F$-statistics \textit{sensu} Patterson are two of the most widely used population genetic tools to study human genetic variation. Here, I derive explicit connections between the two approaches and  show that these two methods are closely related. $F$-statistics have a simple geometrical interpretation in the context of PCA, and orthogonal projections are a key concept to establish this link. I show that negative $F_3$ corresponds to a circle on a PCA plot, and that $F_4$ is related to an angle measurement. I illustrate my results on two examples, one of Western Eurasian, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and that PCA-plots are effective at predicting $F$-statistics. My results extend $F$-statistics to more continuous population models, moving towards a more complete descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa \citep{ramachandran2005}, Austronesian \citep{stoneking2016} or Bantu expansions \citep{li2014} lead to more gradual patterns of genetic diversity over space \citep{francois2010}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{boca2020}.

@@

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, jombart2009, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}. These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations. These analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.

@@

\subsection{Formal Definition of $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci, and for simplicity I will assume that there is no missing data. Let $x_{il}$ denote the frequency of an arbitrary allele at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

\subsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. On trees, all disjoint paths are independent (i.e. orthogonal) from each other, and thus the external branches vanish under the projection.

One issue with the geometric approach of \cite{oteo-garcia2021} is that  each SNP (commonly in the millions) adds a dimension, but high-dimensional spaces are hard to visualize, interpret and analyze. Fortunately, it has been commonly observed that population structure is often quite low-dimensional, and only a  few PCs frequently provide a good approximation of the covariance structure in human genetic variation data \citep{patterson2006}. Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.

@@

\subsubsection{Missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}. One alternative suggested by the theory developed here could be to first perform a PCA to impute missing data, and then calculate $F$-statistics using this imputed data set. Using this approach, the accuracy of $F$-statistics including low-coverage samples would be greatly increased, as they can ``borrow'' information from closely related samples.

@@

An alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}. These associations are not exact, and impacted e.g. by the size and composition of the analyzed data set. On the other hand, we can use $F$-statistics to formally test orthogonality. Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary. Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e. between-population diversity is independent of within-population diversity). This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages or admixture events.
